### 2017-06-29 motivation.tex
motivation

In this section, we motivate our ideas via an example written in
\txnimp - a functional language equipped with a \C{DB} monad to
compose database computations. Besides the usual \C{bind} and
\C{return} combinators, monad offers an \C{atomically} combinator that
executes a database computation and returns the result.

@@

Fig.~\ref{fig:new_order_code} shows a simplified version of the TPC-C
\C{new\_order} transaction written in \txnimp (the \C{do} syntax ala
Haskell is used for convenience). TPC-C is an Online Transaction
Processing (OLTP) benchmark that models an order-processing system for
a wholesale parts supply business. The business logic is captured in 5
database transactions that operate on 9 tables. \C{new\_order} is one
such trasaction that operates on \C{District}, \C{Order},
{New\_order}, \C{Stock}, and \C{Order\_line} tables. The transaction
acts on the behalf of a customer, whose id is \C{c\_id}, to place a
new order for the given set of items (\C{item\_reqs}), to be served by
a warehouse under the district identified by \C{d\_id}. The
transaction does so by invoking appropriate SQL functionality,
captured in Fig.~\ref{fig:new_order_code} by various calls to
functions under the \C{SQL} module. All \C{SQL} functions take the
table name (a nullary constructor) as their first argument. The
higher-order \C{SQL.select1} function accepts a boolean function that
describes the selection criteria, and returns any record that meets
the criteria (it models the SQL query \C{SELECT \ldots\xspace LIMIT
1}). \C{SQL.update} also accepts a boolean function (3rd argument) to
select the records to be updated. Its 2nd argument is a function that
maps each selected record to a new (updated) record. \C{SQL.insert}
inserts a given record into the specified table in the database.

The \C{new\_order} transaction inserts a new \C{Order} record, whose
id is the sequence number of the next order under the given district
(\C{d\_id}). The sequence number is stored in the corresponding
\C{District} record, and updated each time a new order added to the
system. Since each order may request multiple items (\C{item\_reqs}),
an \C{Order\_line} record is created for each requested item to relate
the order with the item. Each item has a corresponding record in the
\C{Stock} table, which keeps track of the quantity of the item left in
stock (\C{s\_qty}). The quantity is updated by the transaction to
reflect the processing of new order (if the stock quantity falls below
10, it is automatically replenished by 91).

TPC-C defines multiple invariants, called \emph{consistency
conditions}, over the state of the application in the database. One
such consistency condition is the requirement that for a given order
\C{o}, the order-line-count field (\C{o.o\_ol\_cnt}) should reflect
the number of order lines under the order, i.e., the number of
\C{Order\_line} records whose \C{ol\_o\_id} is equal to \C{o.o\_id}.
In a sequential exectution, it is more-or-less easy to see how this
condition condition is preserved by the \C{new\_order} transaction by
examining Fig.~\ref{fig:new_order_code}. A new \C{Order} record is
added with its \C{o\_id} distinct from the existing order ids, and its
\C{o\_ol\_cnt} set equal to the size of \C{item\_reqs} set. The
\C{foreach} loop runs once for each \C{item\_req}, adding a new
\C{Order\_line} record for each requested item, with its \C{ol\_o\_id}
field set to \C{o\_id}. Thus, by the end of the loop, the number of
\C{Order\_line} records in the database, whose \C{ol\_o\_id} is equal
to \C{o\_id}, is equal to the size of the \C{item\_req} set, which
inturn is equal to the \C{Order} record's \C{o\_ol\_cnt} field. Thus
the consistency condition is preserved.

While the aforementioned reasoning is simple to perform manually, the
sequential verification of database programs, such as the one in
Fig.~\ref{fig:new_order_code}, is difficult to automate owing to the
presence of loops and collections. Thus verification of serializable
transactions is comparable to the verification of sequential
(imperative) programs in terms of difficulty. It is therefore not
unreasonable to expect the reasoning machinery (i.e., program logics
and inference techniques) developed for imperative programs to be
equally effective on serializable transactions.

Unfortunately, transactions are rarely serializable in practice, and
presence of weak isolation significantly complicates the reasoning
about database programs. Weak isolation allows a transaction to
witness the effects of concurrent transactions during its
execution\footnote{Weak isolation doesn't violate atomicity as long as
the witnessed effects are those of the committed transactions}.
Various levels of weak isolation differ in what effects each allows
the transaction to witness, and when. Admitting interference from
concurrent transactions may violate the invariants that would
otherwise hold under a fully isolated execution.

Let us reconsider the \C{new\_order} transaction, which has been
previously shown to preserve TPC-C's consistency condition under a
serializable execution. However, serializability is rarely the default
isolation level on databases. For instance, the default on Postgres is
Read Committed (RC), which is considerably weaker than
serializability. An RC transaction is isolated from \emph{dirty
writes}, i.e., writes of uncommitted transactions, but is allowed to
witness the writes of concurrent transactions as soon as they are
committed. Thus, with two concurrent instances of the \C{new\_order}
transaction (call them $T_1$ and $T_2$) trying to place new orders for
different customers under the same district (\C{d\_id}), executions
shown in Fig.~\ref{fig:new_order_execs} are possible on Postgres.

@@

The figure depicts an execution as a series of read, write, and commit
operations. In the execution on left, the \C{new\_order} instance
$T_1$ (green) reads the \C{d\_next\_o\_id} field of the district
record for \C{d\_id}, but before it increments the field, another
\C{new\_order} instance ($T_2$) begins its execution and commits. Note
that $T_2$ reads the same \C{d\_next\_o\_id} value as $T_1$, and
inserts new \C{Order} and \C{Order\_line} records with their \C{o\_id}
and \C{ol\_o\_id} fields (resp.) equal to \C{d\_next\_o\_id}. $T_2$
also increments the \C{d\_next\_o\_id} field, which $T_1$ has already
acccessed. This is allowed because reads do not obtain a mutually
exclusive lock on most databases, including Postgres. Post $T_2$'s
commit, $T_1$ resumes execution and adds new \C{Order} and
\C{Order\_line} fields with the same order id as $T_1$. Thus, by the
end of the execution, \C{Order\_line} records inserted by $T_1$ and
$T_2$ all bear the same order id. There are also two \C{Order} records
with the same district id (\C{d\_id}) and order id, none of whose
\C{o\_ol\_cnt} reflects the actual number of \C{Order\_line} records
inserted with that order id. This clearly violates TPC-C's consistency
condition.

Note that, in the above example, there is no Write-Write conflict, or
the \emph{lost update} anomaly that often characterizes concurrency
bugs, such as illegal concurrent withdraws on a bank
account~\cite{poloniexbug}. While $T_1$ and $T_2$ both increment the
\C{d\_next\_o\_id} field of the district record, they do so atomically
(Line 4 of Fig.~\ref{fig:new_order_code}), allowing both updates to be
present in the final state. Likewise, four other well-known anomalies
that characterize RC isolation~\cite{berenson}, namely, \emph{fuzzy
reads}, \emph{phantom reads}, \emph{read skew}, and \emph{write skew},
are also not exhibited by the example. Thus, any program analysis,
static or dynamic, that aims to determine appropriate isolation by
checking for possible manifestations of known anomalies fails to
identify grounds for promoting the isolation level of \C{new\_order}.
Yet, if we take the semantics of the application into account, it is
clear that RC is not an appropriate isolation level for
\C{new\_order}. Clearly, there is a need for a high-level verification
framework that lets us reason about transaction semantics in
conjunction with the semantics of weak isolation.

Facilitating high-level reasoning about weak isolation is however not
easy owing to the low-level nature of its specification, either in
terms of the implementation details~\cite{gray}, allowed
anomalies~\cite{berenson}, or well-formedness condition on program
traces~\cite{adyaphd,gotsmanconcur15}. For instance, does promoting
the isolation level of \C{new\_order} to Repeatable Read (RR) help in
preserving the consistency condition? To answer this question, one has
to relate the low-level specification of RR given in terms of
trace-based relations~\cite{adyaphd,gotsmanconcur15} to the semantics
of TPC-C, which is difficult. Alternatively, one could reason about
how RR is implemented, say on Postgres, and reason about the
\C{new\_order} implementation in conjunction with Postgres's RR
implementation. However, this is not any easier, considering that
Postgres employs a host of techniques, including Multiversion, locks
and rollbacks, to implement RR. The verification problem becomes even
more difficult if the aim is to facilitate automation. Our experience
suggests that a proof framework that requires programmers to relate
trace-based isolation specifications to program semantics necessarily
imposes heavy annotation burden, requires painstakingly long and
verbose proofs, and does not lend itself to automation. Weak isolation
aside, there are also the conventional challenges that face
(automatic) verification of concurrent programs: facilitating modular
proofs, inferring intermediary assertions and loop invariants that are
stable w.r.t interference, etc. A proof system for weakly isolated
transactions has to necessarily address these challenges, along with
those that arise due to weak isolation.

The proof system and the automation framework proposed in this paper
overcomes or circumvents the aforementioned problems. The first
observation that underlies our approach is that, while databases
implement weak isolation employing a host of techniques to facilitate
maximum concurrency despite the isolation, the operational semantics
of such implementations can nonetheless be captured using a simple
abstraction of the database state exposed to the application. For
instance, consider how Postgres executes an RR transaction. A
(conceptual) snapshot of the database state is created at the
beginning of the transaction against which its reads are executed.
When the transaction attempts to update a record, Postgres first
checks if the current version of the record in the database is same as
the snapshot version. If yes, then an exclusive lock is obtained on
the record, and the updatation is carried out. If the record is
already locked, Postgres waits (blocks the transaction) until the lock
is released. Once obtained, an exclusive lock is held until the
transaction commits or rolls back. If the aforementioned version check
fails (i.e., database has a later version than the snapshot), then the
transaction is rolled back and re-executed. Thus, the implementation
of RR on Postgres is fairly involved. Nonetheless, its operational
semantics can be described straightforwardly in terms of the legal
transitions of the database state (denoted, $\Delta \longrightarrow
\Delta'$) allowed while an RR transaction is in progress. Since
transaction's reads are always served from a snapshot, no state
changes are witnessed while the execution is in progress. Thus,
insofar as an RR transaction is concerned, the database state doesn't
change (i.e., $\Delta' = \Delta$) during the execution. Note that the
uncommitted writes of the transaction are not part of $\Delta$;
instead we imagine a transaction-local state ($\delta$) that is
populated with only the uncommitted local writes. When the transaction
commits, the local state ($\delta$) is flushed to the global state
($\Delta$) to yeild a committed state. However, unlike reads, commit
is performed against the current state of the database (call it
$\Delta_c$); not the snapshot. Thus, after the transaction finishes
execution, but before it commits, the transaction witnesses the
effects of all concurrent transactions in a single database state
transition $\Delta \longrightarrow \Delta_c$. Postgres's RR
implementation effectively constrains this transition. Firstly, due to
the version check, the current transaction cannot update a data item
that was already updated by a concurrent transaction. Secondly, due to
exclusive write locks, a data item updated by the current transaction
cannot be overwritten by a concurrent transaction. Putting these two
together, we know that the transition $\Delta \longrightarrow
\Delta_c$ does not involve any concurrent transactions that write to
the same data items as the current RR transaction. Since $\delta$
records the writes of the current transaction, intuitively, forall
$x\in\delta$, $\Delta_c(x) = \Delta(x)$. To summarize, the operational
semantics of Postgres's RR implementation can be captured as
constraints on the transitions of the database state ($\Delta
\longrightarrow \Delta'$) during the lifetime an RR
transaction ($T$):
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
\end{itemize}

@@

>>>>

We have developed and implemented our ideas in an OCaml embedded DSL
that supports a \C{DB} monad to define and compose database
computations given in terms of SQL operations; the monadic \C{do}
syntax is borrowed from Haskell.  Besides the usual \C{bind} and
\C{return} combinators, monad offers an \C{atomically} combinator that
executes a database computation as an atomic transaction and returns
the result.

@@

Fig.~\ref{fig:new_order_code} shows a simplified version of the TPC-C
\C{new\_order} transaction.  TPC-C is a well-known Online Transaction
Processing (OLTP) benchmark that models an order-processing system for
a wholesale parts supply business. The business logic is captured in 5
database transactions that operate on 9 tables. \C{new\_order} is one
such transaction that operates on \C{District}, \C{Order},
\C{New\_order}, \C{Stock}, and \C{Order\_line} tables. The transaction
acts on the behalf of a customer, whose id is \C{c\_id}, to place a
new order for the given set of items (\C{item\_reqs}), to be served by
a warehouse under the district identified by \C{d\_id}. The
transaction does so by invoking appropriate SQL functionality,
captured by various calls to functions under the \C{SQL} module. All
\C{SQL} functions take the table name (a nullary constructor) as their
first argument. The higher-order \C{SQL.select1} function accepts a
boolean function that describes the selection criteria, and returns
any record that meets the criteria (it models the SQL query \C{SELECT
  \ldots\xspace LIMIT 1}). \C{SQL.update} also accepts a Boolean
function (3rd argument) to select the records to be updated. Its
2$%nd$ argument is a function that maps each selected record to a new
(updated) record. \C{SQL.insert} inserts a given record into the
specified table in the database.

The \C{new\_order} transaction inserts a new \C{Order} record, whose
id is the sequence number of the next order under the given district
(\C{d\_id}). The sequence number is stored in the corresponding
\C{District} record, and updated each time a new order is added to the
system. Since each order may request multiple items (\C{item\_reqs}),
an \C{Order\_line} record is created for each requested item to relate
the order with the item. Each item has a corresponding record in the
\C{Stock} table, which keeps track of the quantity of the item left in
stock (\C{s\_qty}). The quantity is updated by the transaction to
reflect the processing of new order (if the stock quantity falls below
10, it is automatically replenished by 91).

TPC-C defines multiple invariants, called \emph{consistency
  conditions}, over the state of the application in the database. One
such consistency condition is the requirement that for a given order
\C{o}, the \emph{order-line-count} field (\C{o.o\_ol\_cnt}) should
reflect the number of order lines under the order, i.e., the number of
\C{Order\_line} records whose \C{ol\_o\_id} is equal to \C{o.o\_id}.
In a sequential execution, it is easy to see how this condition is
preserved.  A new \C{Order} record is added with its \C{o\_id}
distinct from the existing order ids, and its \C{o\_ol\_cnt} is set to
be equal to the size of \C{item\_reqs} set. The \C{foreach} loop runs
once for each \C{item\_req}, adding a new \C{Order\_line} record for
each requested item, with its \C{ol\_o\_id} field set to
\C{o\_id}. Thus, at the end of the loop, the number of \C{Order\_line}
records in the database, whose \C{ol\_o\_id} is equal to \C{o\_id}, is
equal to the size of the \C{item\_req} set, which in turn is equal to
the \C{Order} record's \C{o\_ol\_cnt} field, thus preserving the
consistency condition.

Because the aforementioned reasoning is reasonably simple to perform
manually, it may be feasible to apply techniques for concurrent
program verification~\cite{....} to automate much of this effort.
Serializability aids the tractability of verification by preventing
any interference among concurrently executing transactions.  Under
weak isolation\footnote{Weak isolation doesn't violate atomicity as
  long as the witnessed effects are those of committed transactions},
interferences of various kinds are permitted.  Although various forms
of weak isolation differ in the kinds of effects they allow and when
those effects are made visible, admitting interference of any kind
from concurrent atomic transactions makes verifying program invariants
significantly harder than under a serializable execution.

To illustrate, reconsider the behavior of the \C{new\_order}
transaction when executed with a \emph{Read Committed} (RC) isolation
level, the default isolation level in PostgreSQL, a widely used
open-source database system.  An RC transaction is isolated from
\emph{dirty writes}, i.e., writes of uncommitted transactions, but is
allowed to witness the writes of concurrent transactions as soon as
they are committed. Thus, with two concurrent instances of the
\C{new\_order} transaction (call them $T_1$ and $T_2$), both
concurrently placing new orders for different customers under the same
district (\C{d\_id}), PostgreSQL admits the two executions shown in
Fig.~\ref{fig:new_order_execs}.

@@

The figure depicts an execution as a series of read, write, and commit
operations. In the execution on the left, the \C{new\_order} instance
$T_1$ (green) reads the \C{d\_next\_o\_id} field of the district
record for \C{d\_id}, but before it increments the field, another
\C{new\_order} instance ($T_2$) begins its execution and commits. Note
that $T_2$ reads the same \C{d\_next\_o\_id} value as $T_1$, and
inserts new \C{Order} and \C{Order\_line} records with their \C{o\_id}
and \C{ol\_o\_id} fields (resp.) equal to \C{d\_next\_o\_id}. $T_2$
also increments the \C{d\_next\_o\_id} field, which $T_1$ has already
acccessed. This is allowed because reads do not obtain a mutually
exclusive lock on most databases, including PostgreSQL. After $T_2$'s
commit, $T_1$ resumes execution and adds new \C{Order} and
\C{Order\_line} fields with the same order id as $T_1$. Thus, by the
end of the execution, \C{Order\_line} records inserted by $T_1$ and
$T_2$ all bear the same order id. There are also two \C{Order} records
with the same district id (\C{d\_id}) and order id, none of whose
\C{o\_ol\_cnt} reflects the actual number of \C{Order\_line} records
inserted with that order id.  This clearly violates TPC-C's consistency
condition.

Notably, this example does not exhibit many common concurrency bugs
that arise from incorrect use of weak isolation such as
\emph{write-write} conflicts, or \emph{lost updates}.  While $T_1$ and
$T_2$ both increment the \C{d\_next\_o\_id} field of the district
record, they do so atomically (Line 4 of
Fig.~\ref{fig:new_order_code}), allowing both updates to be present in
the final state. Likewise, other well-known anomalies that
characterize RC isolation~\cite{berenson} such as \emph{fuzzy reads},
\emph{phantom reads}, \emph{read skew}, and \emph{write skew}, are
also not exhibited by the example.  Thus, program analyses that aim to
determine appropriate isolation by checking for possible
manifestations of these anomalies would fail to identify grounds for
promoting the isolation level of \C{new\_order} to something stronger.
Yet, if we take the semantics of the application into account, it is
quite clear that RC is not an appropriate isolation level for
\C{new\_order}. Clearly, the demonstrates the need for a high-level
verification framework that lets us reason about application semantics
in conjunction with the semantics of weak isolation.

While reasoning in terms of anomalies is cumbersome, and as the above
example shows often inadequate, reasoning about weak isolation in
terms of low-level operational details faces similar hurdles given the
conceptual gap between properties defined by consistency conditions
and data structures and state maintained by implementations.  To
overcome these challenges, we develop a framework that axiomatizes the
behavior of weak isolation implementations to enable modular reasoning
using well-established techniques such as Rely-Guarantee~\cite{...}
augmented to deal with weakly-isolated transactions.

To illustrate, consider how we might verify that \C{new_order} is
sound when executed under a \emph{Repeatable Read} (RR) isolation
level.  PostgreSQL executes an RR transaction by taking a
(conceptual) snapshot of the database state before the transaction
begins that is used by reads performing the transaction.  When the
transaction attempts to update a record, a check is made to determine
if the current version of the record in the database is same as the
snapshot version. If it is, an exclusive lock is obtained on the
record, and the update is performed. If the record is already locked,
the transaction blocks until the lock is released.  When the lock is
obtained, is held exclusively until the transaction commits or rolls
back. If the aforementioned version check fails (i.e., the database
has a later version than the snapshot), the transaction is rolled back
and re-executed.

Although this implementation, comprising many thousands of lines code,
is highly involved, its behavior in terms of state transitions on the
database is fairly simple.  Since transaction's reads are always
served from a snapshot, no state changes are witnessed while the
execution is in progress. Thus, insofar as an RR transaction is
concerned, the database state does not change during the execution.
Uncommitted writes are recorded in a transaction-local state.  When
the transaction commits, the local state is atomically written to the
global state to yield a global state that reflects the transaction's
updates.  However, unlike a strongly isolated serializable
transaction, the commit operation is performed against the current
state of the database, not the snapshot. Thus, after the transaction
finishes execution, but before it commits, the transaction is able to
witness the effects of all concurrent transactions.  The PostgreSQL RR
implementation effectively constrains this transition.

We can axiomatize this operational description by observing that (a)
due to the version check, the current transaction cannot update a data
item that was already updated by a concurrent transaction, and (b) due
to the use of exclusive write locks, a data item updated by the
current transaction cannot be overwritten by a concurrent transaction.
If $\Delta$ is the state of the database when an RR transactions
finishes, and $\Delta_c$ is the state visible to the transaction at
the point of commit, we know the transition from $\Delta$ to
$\Delta_c$ (written $\Delta \longrightarrow \Delta_c$) cannot exhibit
effects from any concurrent transactions that write to the same data
items as the current RR transaction.  Similarly, if $\delta$ denotes
a local log that relates transaction variables being written with their updated values,
then $\forall x\in\mathit{dom}(\delta)$, $\Delta_c(x) =
\Delta(x)$. To summarize, the operational semantics of PostgresSQL's RR
implementation can be captured as an axiomatization over transitions of
the database state ($\Delta \longrightarrow \Delta'$) during the
lifetime an RR transaction ($T$):
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
\end{itemize}

This simple characterization of RR isolation allows us to verify the
consistency condtions associated with the \C{new\_order}
transaction. First, since the database doesn't change ($\Delta' =
\Delta$) (where $\Delta'$ represents the state after the transaction
commits), during the execution of the transaction's body, we can
reason about \C{new\_order} as if it executed in complete isolation
until its commit point, leading to a verification process similar to
what would been applied when reasoning about serializability.  When
\C{new\_order} finishes execution and becomes ready to commit, a
transition that transfers the transaction'slocal writes ($\delta$) to
the unchanged database state ($\Delta$).  However, we are required to
consider the interference from concurrent transitions at this point,
which might change the database state from $\Delta$ to $\Delta_c$. If
this interference includes the effects of a concurrent \C{new\_order}
transaction (with the same \C{d\_id}), then verification fails as
described previously (Fig.~\ref{fig:new_order_execs}). Fortunately,
sequential reasoning shows that this is impossible - RR prevents a
concurrent \C{new\_order} transaction that modifies the same
\C{District} record as the current transaction (concretely, since the
record is already present in the current transaction's local log, any
transition from $\Delta$ to $\Delta_c$ cannot change this record).
Applying such axiomatic reasoning on \C{new\_order} allows us to prove
that the TPC-C invariant holds when the transaction is executed under
PostgreSQL's RR isolation.  Our proof framework generalizes this style
of reasoning to various isolation levels on databases.

The second observation that informs our approach is one that pertains
to automation.

@@

For example, \emph{Repeatable
  Read} is a stronger isolation level that RC; we might ask whether
executing the \C{new_order} transaction under this isolation level
would preserve the consistency conditions of the transaction.  To
answer this question, we must relate the low-level specification of RR
given in terms of e.g., trace-based
relations~\cite{adyaphd,gotsmanconcur15} to the semantics of TPC-C,
which is difficult.   Alternatively, one could reason about how RR is
implemented, say on PostgresSQL, and reason about the \C{new\_order}
implementation in conjunction with Postgres's RR
implementation. However, this is not any easier, considering that
Postgres employs a host of techniques, including Multiversion, locks
and rollbacks, to implement RR. The verification problem becomes even
more difficult if the aim is to facilitate automation. Our experience
suggests that a proof framework that requires programmers to relate
trace-based isolation specifications to program semantics necessarily
imposes heavy annotation burden, requires painstakingly long and
verbose proofs, and does not lend itself to automation. Weak isolation
aside, there are also the conventional challenges that face
(automatic) verification of concurrent programs: facilitating modular
proofs, inferring intermediary assertions and loop invariants that are
stable w.r.t interference, etc. A proof system for weakly isolated
transactions has to necessarily address these challenges, along with
those that arise due to weak isolation.
